Variance reduction for MCMC methods via martingale representations

### 2020-01-20 aap/Revision/vr-mcmc-revision.tex
update

\section{Bounds for moments of  ULA}\label{sec:appendix}
For $\gamma>0$, the Markov kernel $R_\gamma$
on $(\mathbb R^d,\mathcal B(\mathbb R^d))$
is defined by
$$
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
$$
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_{\boldsymbol{\gamma}}^{k,l}$ is
$
Q_{\boldsymbol{\gamma}}^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k}.
$
 We make the following assumptions.

@@

Due to \cite[Lemma~14]{brosse:2018:diffusion}, it is known that conditions (H1) and (H2) imply the result

\begin{description}
\item[{\bf (H3)}] There exist $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
\begin{eqnarray*}
\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
\end{eqnarray*}
\end{description}

\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^D$ satisfies conditions {\bf (H1)} and {\bf (H3)} and without loss of generality consider $\nabla U(0) = 0$. Then the kernel \ref{} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{-\left(\frac{m}{2}\right)}$, $C = 3K_1^2 + 2D + C_1$ for some absolute constant $C_1 > 0$, , which does not depend on $\gamma, D, x$.
\end{lem}

\begin{proof} Consider $V(x) = 1 + \|x\|^2$, then
$$
P_{\gamma}V(x) = \int\limits_{\mathbb{R}^d}V(y)P_{\gamma}(x,dy) = \int\limits_{\mathbb{R}^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy = 
$$

$$
=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
$$
Let us first consider the case $x \notin B(0,K_1)$. Note that $\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2$ and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$. It remains to notice that due to {\bf C1} and {\bf C2}
$$
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
$$
$$
 \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
$$
Thus, plugging everything into expression for $P_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
$$
P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + 2\gamma d
$$
Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d\right) \leq 
$$

$$
\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(3K_1^2 + 2D + C_1)
$$
for some absolute constant $C_1$.
\end{proof}

@@

$$
\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{3K_1^2 + 2D + C_1}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
$$
for some $C_1 > 0$, which does not depend on $d, \gamma$ and $n$.
\end{lem}

>>>>

\section{Bounds for moments of  ULA}\label{sec:appendix}
For $\gamma>0$, the Markov kernel $R_\gamma$
on $(\mathbb R^d,\mathcal B(\mathbb R^d))$
is defined by
\begin{equation}
\label{eq:kernel}
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
\end{equation}
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_{\boldsymbol{\gamma}}^{k,l}$ is
$
Q_{\boldsymbol{\gamma}}^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k}.
$
 We make the following assumptions.

@@

Due to \cite[Lemma~14]{brosse:2018:diffusion}, conditions {\bf (H1)} and {\bf (H2)} imply that there exist $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
\begin{equation}
\label{eq:H3}
\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
\end{equation}

\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$. Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}. 
\end{lem}

\begin{proof} Consider $V(x) = 1 + \|x\|^2$, then
$$
R_{\gamma}V(x) = \int\limits_{\mathbb{R}^d}V(y)P_{\gamma}(x,dy) = \int\limits_{\mathbb{R}^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy = 
$$

$$
=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
$$
Let us first consider the case $x \notin B(0,K_1)$. Note that 
$$
\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2
$$ 
and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d)$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$. It remains to notice that due to {\bf H1} and {\bf H2} (which implies \ref{eq:H3}),
$$
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
$$

$$
 \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
$$
Thus, plugging everything into expression for $R_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
$$
R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (2d + m)
$$
Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d\right) \leq 
$$

$$
\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_1^2}{8} + 2d + m)
$$
for some absolute constant $C_1$.
\end{proof}

@@

$$
\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{3K_1^2 + 2D + C_1}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
$$
\end{lem}
